\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage{geometry}
\geometry{margin=1in}
\setstretch{1.1}

\title{Left-Branch Extraction in English: A CGEL-Based Stress Test}
\author{(exploratory note)}
\date{\today}

\begin{document}
\maketitle

\section{Aim}
We examine why English resists left-branch extraction (LBE) from noun phrases when described in the CGEL architecture (\emph{NP} $\rightarrow$ \emph{DetP} in determiner function $+$ \emph{Nom}). Goal: stress-test non-generative explanations; if the account collapses under counterexamples, we discard it. No claim of proof.

\section{CGEL scaffolding}
In CGEL the NP consists of a Determinative Phrase (DetP) plus a nominal (Nom). The nominal contains the head noun with its complements and pre-/post-head modifiers. Possessive/genitives are DetP in determiner function; prenominal adjectives sit inside Nom before the head noun.

\section{Multiple live hypotheses (meant to be broken)}
\begin{enumerate}[label=H\arabic*., leftmargin=1.2cm]
  \item \textbf{Relevance/QUD}: Bare LBE fails because adjectives/determiners are rarely the primary QUD target; pied-piping foregrounds the whole nominal. \emph{Torpedo}: find contexts where the modifier alone is the QUD and bare LBE is still rejected.
  \item \textbf{Prosodic adjacency/edge}: English requires modifier--noun adjacency and lacks a prosodic edge for stranded modifiers; pied-piping keeps adjacency. \emph{Torpedo}: show a prosodically acceptable LBE that speakers like.
  \item \textbf{Recoverability/morpho-cues}: With no case/agreement on modifiers/Det, a stranded noun isn't linkable. \emph{Torpedo}: add overt cues (classifier, agreement play) and see if LBE stays bad.
  \item \textbf{Packaging integrity (CGEL NP as minimal unit)}: English moves or elides DetP+Nom as a package; breaking it is illicit. \emph{Torpedo}: find a productive English construction that leaves an overt noun while extracting its left branch.
  \item \textbf{Linearization cost}: LBE disrupts required order; freer-order languages avoid the cost. \emph{Torpedo}: show an English construction that tolerates similar disruption but still bans LBE.
  \item \textbf{Freezing/movement history}: Prior NP movement freezes interior; baseline LBE is already bad, topicalisation worsens it. \emph{Torpedo}: topicalised DPs where pied-piping is perfect.
  \item \textbf{Usage/frequency}: LBE is rare in input; gradient unacceptability reflects expectations. \emph{Torpedo}: exposure/training that fails to improve LBE.
  \item \textbf{Determiner semantics}: Wh-determiners need an overt domain; stranding the noun breaks presuppositions. \emph{Torpedo}: a determiner whose semantics is satisfied by a stranded noun yet LBE is still bad.
\end{enumerate}

None of these is assumed decisive; keep all on the table until a clear counterexample collapses them.

\section{English data: baseline and probes (for all hypotheses)}
\subsection{Canonical bad LBE}
\begin{enumerate}[label=\alph*)]
  \item *\textbf{How expensive} did they buy \underline{a} car? (adjectival LBE from Nom)
  \item *\textbf{Whose} did you read \underline{book}? (possessor/genitive LBE from DetP)
\end{enumerate}

\subsection{Pied-piping successes}
\begin{enumerate}[label=\alph*)]
  \item \textbf{How expensive a car} did they buy?
  \item \textbf{Whose book} did you read?
\end{enumerate}

\subsection{Prosody and QUD attempts}
\begin{enumerate}[label=\alph*)]
  \item Context QUD: ``Which \emph{color}?'' \\
        *\textbf{Which red} did you buy car? (still crashes)
  \item Contrastive stress: \textbf{HOW} expensive did they buy a car? (judged odd $\rightarrow$ at best marginally improved, not acceptable)
\end{enumerate}

\subsection{Possessor vs adjective asymmetry}
Even in strongly contrastive contexts, possessor LBE seems worse than adjective LBE:
\begin{enumerate}[label=\alph*)]
  \item *\textbf{Whose} did you dent car? (DetP extracted)
  \item ??\textbf{How heavy} did you lift box? (slightly less awful for some speakers)
\end{enumerate}

\subsection{Freezing/topicalisation}
\begin{enumerate}[label=\alph*)]
  \item *That car, \textbf{how expensive a} \underline{\phantom{XX}} did they buy? (pied-piping degraded after topicalisation)
  \item *That car, \textbf{how expensive} did they buy \underline{\phantom{XX}}? (bare LBE still impossible)
\end{enumerate}

\section{Known slippery cases (keep testing)}
\begin{itemize}
  \item \textbf{Echo/reprise}: \emph{You bought WHICH car?} — acceptable but a different construction (does not license information LBE).
  \item \textbf{Sluicing/ellipsis}: \emph{They bought a car, but I don't know how expensive}. Noun is unpronounced; does not show LBE.
  \item \textbf{Fragments/headlines/appositives}: English allows bare nouns or modifiers here; need to check whether any genuinely pronounce the noun and extract the modifier in an information question.
  \item \textbf{Idioms/light nouns}: no solid cases where the adjective alone carries the idiom and could license extraction.
  \item \textbf{Prosody}: no solid reports of contrastive-stress fully rescuing LBE; if found, prosodic-edge accounts gain weight over packaging.
\end{itemize}

\section{What would falsify the account?}
\begin{enumerate}[label=F\arabic*., leftmargin=1.2cm]
  \item Naturally occurring English examples where only the adjective/determiner is extracted in an information question and judged fully acceptable (not echo, not metalinguistic).
  \item Clear evidence that strong contrastive prosody alone makes bare LBE acceptable across speakers.
  \item Contexts where the modifier is the QUD target and bare LBE rates on par with pied-piping.
  \item Any case where possessor LBE patterns like pied-piping in acceptability (would undercut the foreground clash claim for DetP).
\end{enumerate}

\section{Predictions per hypothesis (English)}
\begin{itemize}
  \item \textbf{Relevance/QUD}: modifier-targeted contexts improve LBE slightly but never reach pied-piping; neutral contexts stay bad.
  \item \textbf{Prosodic edge}: contrastive-stress variants should rate better than flat prosody; true rescue would favour this account.
  \item \textbf{Recoverability/morpho-cues}: adding overt cues (e.g., classifier-like of-phrases) should improve LBE; if not, cue hypothesis weakens.
  \item \textbf{Packaging integrity}: no pronounced noun with extracted left branch in ordinary clauses; only pied-piping/ellipsis acceptable.
  \item \textbf{Linearization}: anything that breaks modifier–noun adjacency stays bad; if an adjacency-violating construction is good, this account weakens.
  \item \textbf{Freezing}: topicalised/scrambled DPs degrade even pied-piping; if not, freezing is irrelevant.
  \item \textbf{Usage}: exposure/training might raise ratings marginally; if ratings stay flat, usage is insufficient.
  \item \textbf{Determiner semantics}: determiners that permit domain drop (e.g., \"which ones\") should be better than bare \"which\" + overt noun; if not, semantics isn’t the blocker.
\end{itemize}

\section{Where this could break}
Any robust, context-independent good LBE in English (non-echo, non-ellipsis) sinks packaging/recoverability accounts. Likewise, if prosody or QUD alone fully licenses bare LBE, prosodic/relevance factors outrank packaging. If adjacency-violating constructions are fine, linearization is not the issue. We should be prepared to drop whatever fails first.

\section{Open tasks}
\begin{enumerate}[label=O\arabic*., leftmargin=1.2cm]
  \item Catalogue English near-miss cases (literary, dialect, fragments) and classify them.
  \item Compare bare LBE vs pied-piping across determiners vs adjectives vs quantifiers.
  \item Look for idiom/light-noun cases where the modifier might carry the idiom.
  \item Identify modifier-QUD contexts and check whether pied-piping still wins.
  \item (Crossing to other hypotheses) Note any adjacency-breaking constructions that are grammatical; see if they correlate with better LBE.
\end{enumerate}

\section{Takeaway}
Multiple hypotheses stay on the table; none is claimed decisive. The note is a checklist for where to attack: find genuine English LBE successes or show that prosody/QUD overrides packaging, and we jettison the failing story.

\end{document}
