Below is a publishing roadmap that has worked for other broad‑spectrum programmes (Optimality Theory, Construction Grammar, Exemplar Theory) and that fits the structure of *Grammaticality de‑idealized*.

---

## 0 Orienting facts

| Move                    | Typical vehicle                                                                              | Word limit     | Primary audience              | Goal                             |
| ----------------------- | -------------------------------------------------------------------------------------------- | -------------- | ----------------------------- | -------------------------------- |
| Manifesto               | flagship generalist journal (*Language*, *Glossa*, *Journal of Linguistics*)                 | 9 000–12 000   | all linguists                 | stake the claim, signal ambition |
| Formal skeleton         | theoretically focused journal (*Linguistic Inquiry*, *Natural Language & Linguistic Theory*) | 8 000–10 000   | syntacticians, model builders | show precise machinery           |
| Empirical demonstration | data‑driven journal (*Cognitive Linguistics*, *Frontiers in Language Sciences*)              | 8 000–10 000   | typologists, psycholinguists  | prove the model earns its keep   |
| Tutorial synthesis      | invited handbook chapter, Oxford Research Encyclopedia entry                                 | 7 000–9 000    | newcomers                     | teach the framework              |
| Monograph               | LangSci Press, OUP, CUP                                                                      | 80 000–110 000 | specialists                   | definitive exposition, archival  |

Launching everything at once splinters attention; spacing the pieces 6‑to‑12 months apart keeps the programme in view while letting peer commentary feed forward.

---

## 1 Phase‑by‑phase plan

### Phase A  (three‑month window)

1. Deposit the full draft on Lingbuzz/OSF as a citable preprint.
2. Abstract‑length talk: LSA annual meeting or GLOW.
3. Compact 12‑page proceedings version (LSA, GLOW, or CLS) that previews the model equations (§ formalism) and the decision tree (§ diag‑tree).

> *Purpose:* time‑stamp the idea, collect objections early, and create references you can cite in the later journal papers.

### Phase B  (6–9 months)

**“Manifesto” article**

* Core narrative: §§ 1–3 of the draft (impasse, six components, decision tree).
* One worked example per violation type; strip out most formalism.
* Journal target: *Language* “Target Article” or *Glossa* “General Linguistics”.
* Request open peer commentary if the venue allows; it accelerates uptake.

**Length trimming tricks**

```latex
% in the manifesto file
\input{intro-core.tex}      % 3 800 words
\input{six-components.tex}  % 3 000 words
\input{decision-tree.tex}   % 1 600 words
\input{implications-lite.tex}% 1 800 words
```

A single pass to 10 k words is feasible.

### Phase C  (Year 2)

**Formal‑model article**

* §§ formalism, Fu‑operationalisation, Turkish appendix.
* Emphasise identifiability and the latent‑variable measurement model; park broader philosophical claims.
* Venue: *Natural Language & Linguistic Theory* or *Journal of Language Modelling*.
* Supplementary notebook in an OSF repository that reproduces the simulations.

### Phase D  (Years 2–3)

**Empirical testbeds**

1. **Diachronic S‑curve prediction**
   Corpus study (Penn Parsed Corpora family, ONZE, DARPA YCOE) that estimates Δ > 0 actuation points.
   Outlet: *Language Variation & Change*.

2. **Cross‑linguistic feature integration**
   Gender‑mismatch ratings (Spanish, English, Japanese) + ERP follow‑ups.
   Outlet: *Cognitive Linguistics* or *Second Language Research*.

Each paper can stand alone but cites the manifesto for theory and the model paper for statistics.

### Phase E  (Year 3 or 4)

**Monograph**

*Working title:* *Grammaticality De‑Idealized*.
Publisher choices:

* **LangSci Press** no fees, CC‑BY, typesetting controlled by author (matches your existing workflow).
* **OUP/CUP** wider library penetration, but slower production and costly open‑access option.

Structure:

1. Part I Conceptual foundations (expanded manifesto).
2. Part II Formal model (expanded Phase C).
3. Part III Empirical portfolios (Phases D 1–2 + future studies).
4. Part IV Implications for acquisition, change, clinical work.

---

## 2 Division of labour across pieces

| Content block in current draft                                 | Where it should land                                                        |
| -------------------------------------------------------------- | --------------------------------------------------------------------------- |
| § previous‑impasse                                             | manifesto                                                                   |
| §§ framework core + diag‑tree                                  | manifesto                                                                   |
| § formalism (minus derivations)                                | model article                                                               |
| Full derivations, parameter‑fitting, Turkish harmony extension | online appendix + model article                                             |
| § patterns/ungrammaticality                                    | manifesto (short) + handbook chapter (long)                                 |
| § subjective feeling, misattribution                           | model article                                                               |
| § change‑mechanisms                                            | diachronic empirical paper                                                  |
| § predictions on L2                                            | separate SLA article                                                        |
| Entire bibliography                                            | maintained centrally; each spin‑off pulls required entries via `refsection` |

---

## 3 Conventional launch trajectories in linguistics

1. **Conference sketch → flagship journal article → monograph**
   Optimality Theory (Prince & Smolensky 1991 Rutgers technical report → *LI* 1993 squib → 2004 monograph).

2. **Data‑heavy pilot → theory article → follow‑up series**
   Construction Grammar (Fillmore 1988 proceedings → *Language* 1989 case study → Goldberg 1995 book).

3. **Broad manifesto with commentary → model paper → cross‑field adoption**
   Exemplar phonology (Pierrehumbert 2001 *Frequency & the Emergence of Linguistic Structure* → Goldinger 2007 *Psychonomic Bulletin & Review* → Johnson 2015 monograph).

Your proposed path mirrors trajectory 3: stake a claim, specify the machinery, then let domains test it.

---

## 4 Venue selection rationale

* **High‑impact generalist journals** maximise visibility and force the most stringent review; perfect for the manifesto.
* **Theoretical journals** value formal explicitness; ideal for the equation‑heavy core.
* **Empirical/psycholinguistic journals** reward novel predictions and methods; good for evidence papers.
* **Open‑access monograph** cements the programme and guarantees longevity; LangSci’s workflow dovetails with your LaTeX toolchain and CC‑BY ethos.

---

## 5 Practicalities

* Use the same title stem across pieces (*Grammaticality de‑idealized I, II,* etc.) to make citation chaining obvious.
* Register ORCID‑based author IDs in all deposits; link code/data via DOI (Zenodo or OSF).
* Keep a single `.bib` masterfile; export sectional `refsections` to avoid duplicated entries.
* For each spin‑off, include a boxed “What’s new here” paragraph to head off redundancy complaints.
* Accept that journal review will force trimming; keep full derivations in an online appendix, not in print.

---

## 6 Risks and mitigations

| Risk                                     | Why it arises            | Mitigation                                                                    |
| ---------------------------------------- | ------------------------ | ----------------------------------------------------------------------------- |
| Reviewer sees overlap across submissions | synchronous pipelines    | stagger submissions by ≥ 3 months; cross‑cite explicitly                      |
| Theory dismissed as “all talk”           | manifesto too conceptual | release simulation notebook with manifesto even if details wait for Phase C   |
| Fatigue before monograph finishes        | serial deadlines         | treat monograph as edited compilation of earlier articles; recycle but update |

---

## 7 Suggested 36‑month timeline (indicative)

| Month | Deliverable                         |
| ----- | ----------------------------------- |
| 1     | Lingbuzz preprint                   |
| 2     | LSA abstract                        |
| 6     | proceedings paper                   |
| 9     | manifesto submitted                 |
| 15    | formal model submitted              |
| 18    | gender‑ERP experiment preregistered |
| 24    | diachronic corpus paper submitted   |
| 28    | handbook chapter drafted            |
| 30    | gender‑ERP paper submitted          |
| 36    | monograph manuscript delivered      |

---

### One‑paragraph LaTeX skeleton for the manifesto

```latex
\documentclass{langsci}
\title{Grammaticality de‑idealized: A community‑anchored view}
\author{Brett Reynolds}
\begin{document}
\maketitle

\section{Introduction}
% 600 words

\section{Six mechanisms of (un)grammaticality}
% condensed § framework

\section{Decision tree for quick diagnosis}
% 1 page figure + prose

\section{Implications}
% typology, change, acquisition (one subsection each)

\section{Outlook}
% invite commentary

\bibliography{gdi-manifesto}
\end{document}
```

Compile to stay under 30 printed pages.

---

## Take‑home

1. **Lead with a compact manifesto** in a top‑tier open‑access journal to spark debate.
2. **Follow with a formal model piece** that nails the mathematics and measurement.
3. **Publish at least two data‑driven tests** that any critic must engage.
4. **Consolidate** everything, plus accumulated feedback, in an open‑access monograph.

Spacing the outputs keeps the field talking, gives you natural conference material each year, and builds a citation trail that converges on the book.
